Higher Moments Distances from Robust Subspace for Botnet Detection

### 2019-04-28 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
working on introduction

\title{Robust Subspace and Higher Moments Distances for Botnet Detection}

@@

Anomaly detection is the identification of rare and suspicions events by differing the normal or majority of the data. Anomalies are also referred to as outliers, novelties, noise, deviations or exceptions, and  can be related to network attack, fraud, defect or medical condiction, for examples \cite{bhuyan2014network,ahmed2016survey}. Anomalies can be hard to identify and separate from normal data due to the rare occurrencies of anomalies in comparison to normal events, therefore anomaly detection algorithms have to be robust, high discriminative and able to deal with the imbalanced data problem \cite{he2008learning}.

Several efforts and reasearches aim to avoid network attacks based on known attackers, fingerprints or other patterns. However, distributed attacks organized by botnet has increasing and demanding the development of counter measures in oder to detect and avoid unkown attacks or even to deal with adversarial changes of behavior, location and patterns, in order to avoid detection by behavioral or pattern based detecion systems \cite{gu2008botminer, garcia2014empirical,khattak2015botflex,acarali2016survey,wang2017botnet,Wang2018ddosbotnetssurvey}. To face this problem, it is possible to adopt unsupervised or semi-pervised approches, where it is not necessary known anomalies for trannning models \cite{moustafa2019holistic}.

Robust statistics produce methods that are not unduly affected by outliers.

Robust statistical methods have been developed for estimating location, scale, and regression parameters.

Robust estimates from contaminated data can denoise, remove anomalies or even separate them.

Anomaly detection based on robust estimates has been adopted for network anomaly detection, due to good results and its adaptability.

The distance between robust estimates and new observations can reveal anomalies.

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples by, considering the correlations between the variables and their mean values.

\textbf{MD has been used for distance based anomaly detection with robust estimates} in many areas, and is commonly combined to Fast Minimum Covariance Determinant (Fast-MCD) algorithm \cite{rousseeuw1999fastmcd} for robust estimates.

Network anomaly detection problems are usually characterized by skewed and imbalanced data \cite{Phua2004minority,he2008learning,benson2010network}. Learning algorithms for imbalanced data has been a challenging research topic. The fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}. He and Garcia \cite{he2008learning} present a classification of imbalanced learning methods as sampling, cost-sensitive learning, kernel-based, and active learning. Some techniques can make the data discriminative for better discriminative analytics: Data engineering (i.e. sampling \cite{he2008learning,gu2008botminer}, aggregation \cite{lakhina2005mining,gu2008botminer,callegari2011novel, garcia2014empirical, acarali2016survey}); Feature extraction (i.e. decomposition \cite{vaswani2018robust}, robust estimates \cite{zhou2017anomaly}). To face a high imbalanced data set, we adopt a sampling method based on data aggregation for discriminative feature generation, considering that data aggregation is a widely adoted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. For more information about learning methods from imbalanced data, we refer to He and Garcia \cite{he2008learning}.

\cite{lakhina2005mining} (Mining anomalies using traffic feature distributions).  -Argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Using entropy as a summarization tool, we show that the analysis of feature distributions leads to significant advances on two fronts: (1) it enables highly sensitive detection of a wide range of anomalies, augmenting detections by volume-based methods, and (2) it enables automatic classification of anomalies via unsupervised learning. Anomaly diagnosis framework, comprising both an extension of the subspace method [23] to accommodate multiple data types, and an unsupervised classification technique using simple clustering algorithms. \textbf{To explain feature selection and aggregation}. Goldberg and Shan \cite{goldberg2015importance} argue that processing monitored signals into features is a more fruitful way to build anomaly detection systems than devising ever more complex statistical tests, based on experience and  on putblished real-life examples of anomaly detectors in use at eBay.

% Say that the data is lognormal \cite{benson2010network} and link it to skewness https://www.weibull.com/hotwire/issue47/relbasics47.htm
Some approaches assume a gaussian distribute data, hoever it is not true in real world for many types of network traffic, wich usually is more related to skewed and heavy-tailed distributions. Benson \emph{et al} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highligh that the skeweness and heavy-tailed distributions can impact algorithms that assum a guassian distributed data, as well as can indicate characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis.

Skewness and the kurtosis tests are very sensitive outlier detectors. The properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

PCA is very sensitive to anomalies, thus outliers can corrupt the entire data or can be explored to highlight unespected changes and indicate attacks or frauds \cite{callegari2011novel,Lee2013,vieira2017model}. However, PCA is not precise to reveal flows, time or detailed information for each anomlies, and sometimes relies into visual analysis for pricipal component selection.

We propose a network anomaly detection approach based on robust moment estimates and distance analysis.

Considering that Fast-MCD is a widely adopted robust estimator of location and scatter for anomaly detection, we propose to extend Fast-MCD for computing robust skewness and kurtosis.

We also propose to use Mahalanobis Distance for detecting anomalies based on robust skewness and kurtosis, instead of the location-based usage of classical approaches.

Considering that it is hard to have labeled data in real problems \cite{osanaiye2016distributed}, such as for network attacks or frauds, also considering that known anomalies are constantly changing due to adversarial models or due to lack of regular patterns, we propose a Semi-Supervised approach based on robust moments for network anomaly detection.

We believe that the distribution of anomalous and normal traffic are skewed but with differences that permit separate anomalous of normal traffic. (Assumption and hitotesis). We also believe that aggregated traffic and new features can make the data more discriminative for anomaly detection algorithms. We algo believe that normal traffic presents significant distance to anomalous traffic, which permit a distance based sepparation, that can be augmented by means of robust estimation of normal traffic, that can discard noise or irrelevant outliers and can highlight higher outliers when comparing to new observations

In the context of anomaly-based schemes, this work proposes a statistical approach based on cumulants and Mimimum Covariance Determinant (MCD) for network attack detection. To the best of our knowledge there is no similar model in the literature. We evaluate the accuracy of the proposed framework applied to CTU-13 dataset ??, which is a well known network traffic dataset.

Our approach train from normal traffic and have significant computational cost during training. For prediction, the cost is basically the cost for Mahalanobis distance computation

@@

@@

>>>>

\title{Higher Moments Distances from Robust Subspace for Botnet Detection}

@@

Anomaly detection can be defined as the identification of rare and suspicions events by differing the normal or majority of the data. Anomalies are also referred to as outliers, novelties, noise or deviations, and can be related to network attacks, frauds or defects \cite{bhuyan2014network,ahmed2016survey}. Anomalies can be hard to identify and separate from normal data due to the rare occurrencies of anomalies in comparison to normal events, therefore anomaly detection algorithms have to be high discriminative, robust to contamination and able to deal with the imbalanced data problem \cite{he2008learning}.

Several efforts and reasearches aim to avoid network attacks based on known attackers, fingerprints or other patterns. However, distributed attacks organized by botnet has increasing and demanding the development of counter measures in oder to detect and avoid unkown attacks or even to deal with adversarial changes on behavior, location and other patterns, in order to avoid detection by behavioral or pattern based detecion systems \cite{gu2008botminer, garcia2014empirical,khattak2015botflex,acarali2016survey,wang2017botnet,Wang2018ddosbotnetssurvey}. To face this problem, it is possible to adopt unsupervised or semi-pervised approaches for network anomaly detection, where it is not necessary known anomalies for trannning models \cite{moustafa2019holistic}.

Network anomaly detection problems are usually characterized by skewed and imbalanced data \cite{Phua2004minority,he2008learning,benson2010network}. Learning algorithms for imbalanced data has been a challenging research topic, considering that the fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}. Some techniques can make the data discriminative for better results on imbalanced data: Data engineering (i.e. sampling \cite{he2008learning,gu2008botminer}, aggregation \cite{lakhina2005mining,gu2008botminer,callegari2011novel, garcia2014empirical, acarali2016survey}); Feature extraction (i.e. decomposition \cite{vaswani2018robust}, robust estimates \cite{zhou2017anomaly}).

% He and Garcia \cite{he2008learning} present a classification of imbalanced learning methods as sampling, cost-sensitive learning, kernel-based, and active learning. For more information about learning methods from imbalanced data, we refer to He and Garcia \cite{he2008learning}.

% Say that the data is lognormal \cite{benson2010network} and link it to skewness https://www.weibull.com/hotwire/issue47/relbasics47.htm
Some approaches or widely adopted algorithms assume a gaussian distributed data, however this assumption may not be observed in real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}. Findings of Benson \emph{et al}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highligh that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis. Therefore, learning methods for imbalanced and skewed data have been atracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and kurtosis are higher moments that their measurement can be very sensitive for outlier detection, considering that the properties of the kurtosis has been used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

Principal Component Analysis (PCA) is very sensitive to anomalies, thus outliers can corrupt the entire data or can be explored to highlight unespected changes and indicate attacks or frauds \cite{callegari2011novel,Lee2013,vieira2017model}. However, PCA is not precise to reveal flows, time or detailed information for each anomlies, and sometimes relies into visual analysis for pricipal component selection. However, Robust Principal Component Analysis (RPCA) \cite{candes2011robust} aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix, $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix, $\textbf{S}$, which refers to outliers’ matrix. Therefore, robust subspace learning has been adopted for network anomlay detection \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples by, considering the correlations between the variables and their mean values. MD has been used for distance based anomaly detection with robust estimates in many areas, considering that the distance between robust estimates and new observations can reveal anomalies.

We believe that the skewness of anomalous and normal traffic can reveal characteristcs for improving anomaly detection in imbalanced data. The botnet detection algorithms can also be improved by feature generation from aggregated traffic, considering that new features can make the data more discriminative for imbalanced learning problems. We also believe that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for botnet detection. Therefore, we propose an approach based on higher moments distances from robust subspace for botnet detection.

Considering that it is hard to have labeled data in real problems \cite{osanaiye2016distributed}, such as for network attacks or frauds, also considering that known anomalies are constantly changing due to adversarial models or due to lack of regular patterns, we propose a Semi-Supervised approach based on robust moments for network anomaly detection. We also propose to use Mahalanobis Distance for detecting anomalies based on robust skewness and kurtosis, instead of the location-based usage of classical approaches. Our approach train from normal traffic and have significant computational cost during training. For prediction, the cost is basically the cost for Mahalanobis distance computation To face the anomaly detection from a high imbalanced data set, we adopt a sampling method based on data aggregation for discriminative feature generation, considering that data aggregation is a widely adoted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. We evaluate the accuracy of the proposed framework applied to CTU-13 dataset ??, which is a well known network traffic dataset.

To the best of our knowledge there is no similar model in the literature.

@@

Lakhina \emph{et al} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Using entropy as a summarization tool, we show that the analysis of feature distributions leads to significant advances on two fronts: (1) it enables highly sensitive detection of a wide range of anomalies, augmenting detections by volume-based methods, and (2) it enables automatic classification of anomalies via unsupervised learning. Anomaly diagnosis framework, comprising both an extension of the subspace method [23] to accommodate multiple data types, and an unsupervised classification technique using simple clustering algorithms. \textbf{To explain feature selection and aggregation}. Goldberg and Shan \cite{goldberg2015importance} argue that processing monitored signals into features is a more fruitful way to build anomaly detection systems than devising ever more complex statistical tests, based on experience and  on putblished real-life examples of anomaly detectors in use at eBay.

@@

We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for a robust subspace learning

Benson \emph{et al} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highligh that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis. Therefore, learning methods for imbalanced and skewed data have been atracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and the kurtosis are very sensitive outlier detectors, considering that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

In contrast to classical PCA, our robust PCA detector does not require having a perfect ground-truth for training; it just trains from the (eventually contaminated) background traffic. It is, in fact, an unsupervised learning algorithm. Moreover, the addition of feature selection allows matching the specificities of the traffic the detector is operating on, which can vary widely both in terms of the most frequent attacks and the relative weight of licit applications. Feature selection does require training with a labeled dataset bu

aims to reduce the sensitivity of PCA to outliers, by means of . In particular, RPCA allows for the careful teasing apart of sparse outliers so that the remaining low-dimensional approximation is faithful to the noise-free low-dimensional subspace describing the bulk of the raw data [6, 7, 20]. Specifically, RPCA splits a data matrix X into a low-rank matrix L and a sparse matrix S such that
X = L + S.
